---
title: "440 Case Study I Interim Report"
author: "Jake Epstein, Daniel Spottiswood, Michael Tan, Sahil Patel, Man-Lin Hsiao"
date: "9/10/2019"
output:
  html_document: default
  pdf_document:
    dev: png
    fig_height: 2.5
    fig_width: 4
---

## Introduction

In this report, we will be investigating the relationship between birthweight and various factors. The data we will be analyzing includes children born in North Carolina between 2011 and 2016, totaling 720,000 observations. In addition to birthweight, this data includes factors such as sex, gestation period, smoking behavior, mother's race, and county of residence, which we used as a proxy for mother's socioeconomic status. We will first explore each of these variables and determine which factors are relevant to birthweight, and then use the relevant factors to create and evaluate a linear regression model. We will also use M-estimation to create a more robust model. Finally, we will use cross-validation to compare our chosen linear model with the more robust version.


```{r, message=FALSE, echo=FALSE, warning=FALSE}
## load packages

library(dplyr)
library(ggplot2)
library(MASS)
library(gridExtra)
library(quantreg)
knitr::opts_chunk$set(warning=FALSE, echo=FALSE)
```

```{r}
## read in data
births = read.csv("data/Yr1116Birth.csv", na.strings = "9999")
deaths = read.csv("data/Yr1116Death.csv")
 
## rewrite NAs
births$SEX[which(births$SEX == 9)] = NA
births$CIGPN[which(births$CIGPN == 99)] = NA
births$CIGFN[which(births$CIGFN == 99)] = NA
births$CIGSN[which(births$CIGSN == 99)] = NA
births$CIGLN[which(births$CIGLN == 99)] = NA
births$PARITY[which(births$PARITY == 99)] = NA
births$PLUR[which(births$PLUR == 99)] = NA
births$GEST[which(births$GEST == 99)] = NA
births$MAGE[which(births$MAGE == 99)] = NA
select = dplyr::select
```


## Exploratory Data Analysis

### Birthweight
```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
ggplot(births, aes(x= BWTG))+
  geom_histogram(binwidth=100, colour="black", 
                          aes(y=..density.., fill=..count..), position = "stack") + 
  stat_function(fun = dnorm, color = "red", args = list(mean = mean(births$BWTG, na.rm = TRUE), sd = sd(births$BWTG, na.rm = TRUE))) +
  theme_minimal()
```
</br>
</br>
</br>
Birthweight is close to normally distrubted, with a slight left skew, centered around ~3,300g with a standard deviation of 600g. There appear to be no large outliers in terms of birthweight. 430 birth weights are missing. We see that the left tail is much larger than we would expect in a normal distribution. We'll keep this in mind when evaluating the assumptions of linear regression.

### Sex

```{r}
births$SEX[which(births$SEX == 1)] = "Male"
births$SEX[which(births$SEX == 2)] = "Female"


ggplot(data = na.omit(births), mapping = aes(x = as.factor(SEX), y = BWTG)) +
  geom_boxplot() + xlab("Sex") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + theme_minimal()

births %>%
  na.omit() %>%
  group_by(SEX) %>%
  summarize("Average Birthweight" = mean(BWTG))
```
</br>
</br>
On average, males in our data are about 100g heavier than females at birth.


### Smoking
```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
births = births %>% mutate(
  CIGPN_binary = CIGPN>0,
  CIGFN_binary = CIGFN>0,
  CIGSN_binary = CIGSN>0,
  CIGLN_binary = CIGLN>0
)

births = births %>%
  mutate(total_smoked = CIGFN+CIGSN+CIGLN) %>%
  mutate(smoked_during = total_smoked >0)


ggplot(births %>% filter(smoked_during), aes(x = total_smoked))+
  geom_histogram() +
  xlab("Cigarette Consumption per Week")+
  theme_minimal()

#mean(births %>% filter(smoked_during) %>% select(total_smoked) %>% unlist())

ggplot(births, aes(x= smoked_during, y = BWTG)) +
  geom_boxplot()+
  xlab("Smoked During Pregnancy")+
  ylab("Birth Weight (g)")+
  ggtitle("NC Births, 2011-2016") +
  theme_minimal()

births = births %>%
  mutate(smoking_type = ifelse(smoked_during, 
      ifelse(CIGPN_binary, "before and during", "during only"),
      ifelse(CIGPN_binary, "before only", "none"))) %>%
  mutate(smoking_type = relevel(as.factor(smoking_type), ref = 4))


births %>%
  na.omit() %>%
  group_by(smoking_type) %>%
  summarize("Average Birthweight" = mean(BWTG))

ggplot(births, aes(x= smoking_type, y = BWTG)) +
  geom_boxplot()+
  xlab("Smoking in Relation to Pregnancy")+
  ylab("Birth Weight (g)")+
  ggtitle("NC Births, 2011-2016") +
  theme_minimal()

```
</br>
</br>
Because the average cigarette consumption in our data varied quite a bit, and because this data was likely self-reported, we chose to simplify our analysis on smoking and chose to turn smoking into a categorical variable, with mothers classified as either 1) non-smokers, 2) those who only smoked before pregnancy, 3) those who smoked during but not before pregnancy, and 4) those who smoked throughout.
</br>
</br>
Around 13% of women smoked in the three months leading up to pregnancy and around 10% of women smoked at any point during their pregnancy. Among those who did smoke during pregnancy, the average number of cigarettes smoked during pregnancy was 23. The birthweight of children of mothers who smoked before and during pregnancy was significantly lower than that of the children of nonsmokers, with an average difference of about 200 grams. There is also a significant relationship between birthweight and smoking before pregnancy, even for those who did not smoke during pregnancy.


### Parity

```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
# Check the parity frequencies
births$PARITY = as.numeric(births$PARITY)
births = births %>% 
  mutate(PARITY_truncated = ifelse(
    PARITY > 4, "5+", PARITY)
  )
  
births$PARITY = as.factor(births$PARITY)
ggplot(data = births, mapping = aes(x = PARITY, y = BWTG)) +
  geom_boxplot() + xlab("Parity") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + theme_minimal()

ggplot(births, aes(x = PARITY)) +
  stat_count()+
  theme_minimal()

# Rerun EDA with truncated data
births$PARITY_truncated = as.factor(births$PARITY_truncated)
ggplot(births, aes(x = PARITY_truncated)) +
  stat_count() +
  theme_minimal()

ggplot(data = births, mapping = aes(x = PARITY_truncated, y = BWTG)) +
  geom_boxplot() + xlab("Parity Truncated") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") +
  theme_minimal()

```
</br>
</br>
Independent of other variables, we see a negative relationship between parity and birth weight past the first child. The frequency of parity decreases in an exponential fashion. A second variable was created that truncates parities of at least five to improve interprability and prevent overfitting. The quantity of missing data is relatively small, as there are very few instances of mothers having more than 5 children.

### Plurality

```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
# Check the plurality frequencies
births$PLUR = as.numeric(births$PLUR)

births = births %>% 
  mutate(PLUR_truncated = ifelse(PLUR > 2, "3+", PLUR))

births$PLUR = as.factor(births$PLUR)
ggplot(data = births, mapping = aes(x = PLUR, y = BWTG)) +
  geom_boxplot() +
  xlab("Plurality") +
  ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + 
  theme_minimal()

ggplot(births, aes(x = PLUR)) +
  stat_count() +
  xlab("Plurality") +
  theme_minimal()

# Rerun EDA with truncated data
births$PLUR_truncated = as.factor(births$PLUR_truncated)

ggplot(data = births, mapping = aes(x = PLUR_truncated, y = BWTG)) +
  geom_boxplot() + 
  xlab("Plurality Truncated") + 
  ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + 
  theme_minimal()


ggplot(births, aes(x = PLUR_truncated)) +
  stat_count() + 
  xlab("Plurality") +
  theme_minimal()
```
</br>
</br>
We see a strong non-linear negative relationship between plurality and birth weight. The frequency of pluralities above two is extremely small, and we again see a proportionally small amount of missing data (very few mothers have triplets or more). A second variable was created that truncates pluralities of at least three to improve interprability and prevent overfitting. 

### Gestation

```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}

ggplot(data = births, mapping = aes(x = as.factor(GEST), y = BWTG)) +
  xlab("Gestation Period (weeks)") + 
  ylab("Birth weight (g)") + 
  geom_boxplot() +
  ggtitle("NC Births, 2011-2016") + 
  theme_minimal()


ggplot(births, aes(x = GEST)) +
  stat_count() + 
  xlab("Gestation Period (weeks)") +
  theme_minimal()
```
</br>
</br>
There appears to be a non-linear positive relationship between gestation period and birth weight. The mean gestational period is approximately 38.5 weeks and the period with the highest median weight is 42 weeks. The frequency distribution is left skewed with the majority of babies having a gestational period between 38 and 40 weeks. There is some concern that more extreme gestational periods may lead to higher variance, and it should be noted that there is a chunk of data points with gestational periods of 17 to 21 weeks that have much higher than expected birth weights. There is an extreme outlier with gestational age of 83 weeks. Given that this data point was probably incorrectly recorded (it is biologically nearly impossible for a human to gestate for 83 weeks), we will exclude it from our analysis when building the model.

### Age of Mother

```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
ggplot(data = births, mapping = aes(x = MAGE, y = BWTG)) +
  xlab("Age of Mother (years)") + 
  ylab("Birth weight (g)") + 
  geom_point() +
  geom_smooth(method='lm', na.rm = TRUE) + 
  ggtitle("NC Births, 2011-2016") + 
  theme_minimal()


ggplot(births, aes(x = MAGE)) +
  geom_histogram(binwidth=1, colour="black", 
                          aes(y=..density.., fill=..count..), position = "stack") + 
  stat_function(fun = dnorm, color = "red", args = list(mean = mean(births$MAGE, na.rm = TRUE), sd = sd(births$MAGE, na.rm = TRUE))) +
  xlab("Age of Mother (years)") + 
  theme_minimal()
```
</br>
</br>
Mother's age seems to be fairly normally distributed with a mean of 27.7, with a slightly shorter left tail. There appears to be a positive relationship between the age of the mother and the birth weight. There is no evidence to suggest that the birth weight variance is not constant across the mother's age.

### Race of Mother

```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
births$MRACER = as.factor(births$MRACER)
ggplot(data = births, mapping = aes(x = MRACER, y = BWTG)) +
  xlab("Race of Mother") + 
  ylab("Birth weight (g)") + 
  geom_boxplot() +
  ggtitle("NC Births, 2011-2016") + 
  theme_minimal()

ggplot(births, aes(x = MRACER)) +
  stat_count() +
  xlab("Race of Mother") +
  theme_minimal()
```
</br>
</br>

0 - Other non-White
</br>
</br>
1 - White
</br>
2 - Black or African American
</br>
3 - American Indian or Alaska Native
</br>
4 - Chinese
</br>
5 - Japanese
</br>
6 - Native Hawaiian
</br>
7 - Filipino
</br>
8 - Other Asian 
</br>
</br>

There are significant differences between the average birth weights of mothers of different races. We see that mothers that self-identified as white have the largest mean baby weight at 3.33 kg, while black mothers have the lowest mean baby weight at only 3.07 kg. 58 percent of mothers identify as white, 24 percent identify as black, 12 percent identify as other non-white, and 3 percent identify as other asian.

### County / Socioeconomic Status
```{r, echo=FALSE, message=FALSE, fig.width=5, fig.height=2.5}

#calculate infant mortality by county, to use as a proxy for socioconomic status
deaths_by_county = deaths %>%
  group_by(cores) %>%
  summarize(n_deaths = n()) %>%
  rename(CORES=cores)

births_by_county = births %>%
  group_by(CORES) %>%
  summarize(n_births = n())

infant_mortality = merge(deaths_by_county, births_by_county, by = "CORES") %>%
  mutate(mortality = n_deaths/n_births*100) %>% #note: mortality rate is as percentage
  select(-n_births, -n_deaths)

births = merge(births, infant_mortality, by = "CORES")
bwtg_vs_mortality = lm(data = births, BWTG~mortality)
ggplot(births, aes(x= mortality))+
  geom_histogram() +
  xlab("Mortality Rate (%)")+
  theme_minimal() +
  ggtitle("Histogram of Infant Mortality Rate by County")

ggplot(births, aes(x = mortality, y = BWTG))+
  geom_point() +
  geom_abline(slope = bwtg_vs_mortality$coefficients[[2]], intercept = bwtg_vs_mortality$coefficients[[1]], color = "blue")+
  xlab("Mortality Rate (%)") +
  ylab("Birth Weight (g)") +
  theme_minimal()
```
</br>
</br>
We chose to use infant mortality rate of birth county as a proxy for socioeconomic status, calculated as number of deaths before the age of 1 divided by total number of births in a county. The median county in the data had a infant mortality rate of 0.7%, with the range of infant mortality rates in our dataset ranging from 0.12% to 1.76%. Infant mortality rate of birth county and birth weight appear to have a weak negative linear relationship, and in isolation, a 1 percentage point increase in infant mortality rate is associated with a 157g decrease in expected birth weight.

## Model Building and Selection

### Model 1: Initial Approach
```{r, fig.width=7, fig.height=5}
births_excl = na.omit(births)
births_excl = births_excl[which(births_excl$GEST < 80), ]
births_excl = births_excl %>%
  mutate(GEST2 = GEST^2, GEST3 = GEST ^ 3, GEST4 = GEST^4)

model1 = lm(data = births_excl, BWTG ~ SEX + GEST + PARITY_truncated + PLUR_truncated + smoking_type + MAGE + MRACER + mortality, na.action = "na.exclude")
summary(model1)
par(mfrow = c(2,2))
plot(model1)
plot(births_excl$GEST, model1$residuals, main = "Gestation vs Residuals")
plot(births_excl$PLUR_truncated, model1$residuals, main = "Plurality vs Residuals")
plot(births_excl$PARITY_truncated, model1$residuals, main = "Parity vs Residuals")
plot(births_excl$smoking_type, model1$residuals, main = "Smoking vs Residuals")
plot(births_excl$MRACER, model1$residuals, main = "Raace vs Residuals")
plot(births_excl$MAGE, model1$residuals, main = "Age vs Residuals")
```
</br>
</br>

In building the model, we chose to drop the point at 80 weeks of gestation, which is likely an outlier that has no reason to be there, as no human can possibly gestate for 80 weeks (~1.54 years). Additionally, we dropped all rows with missing data, as the quality of that data may be questionable. We chose to include all factors we explored earlier into this model, and found each of the factors to be statistically significant predictors of birthweight.

The residuals vs fitted values and residuals vs gestational period plot slope downwards, indicating that there is a departure from linearity. A transformation may be helpful. The model may improve if a square term is added for gestation. Furthermore, the Q-Q plot indicates a departure from normality, likely as a result of the left skew of birthweight. We will keep this in mind as we make improvements to this initial model.
</br>
</br>
The residual graph for Plurality (truncated) has decreasing residuals (in terms of absolute value) as plurality increases. This makes sense, as birth weight should get smaller (and as a result range of birth weights should get tighter, leading to smaller absolute value residuals) as more babies share a womb and share nutrients -- More sharing will biologically cause them to come out smaller.
</br>
</br>
The residual graph for Parity (truncated) has pretty random residuals that are all around the same size for each group.
</br>
</br>
The residual graph for Smoking has higher residuals for no smoking than for smoking of any kind. This makes sense, as birth weight could biologically get smaller in the presence of smoking, as smoking can be damaging to the fetus and be detrimental to its growth and weight. This would lead to the range of birth weights of smoking mothers getting tighter, leading to smaller absolute value residuals.
</br>
</br>
The residual graph for Mother's race indicates that residuals are lower for for races 3, 4, 5, 6, and 7 and higher for the other races. This could be something to explore.
</br>
</br>
The residual graph for Mother's age is fairly random, with residuals getting a bit smaller near the beginning and end (<20 years old and >45 years old).


### Model 2: Quadratic Term Added
```{r, fig.width=7, fig.height=5}
model2 = lm(data = births_excl, BWTG ~ SEX + GEST + GEST2 + PARITY_truncated + PLUR_truncated + smoking_type + MAGE + MRACER + mortality, na.action = "na.exclude")
summary(model2)
par(mfrow = c(2,2))
plot(model2)
# plot(model2$fitted.values, model2$residuals)
plot(births_excl$GEST, model2$residuals, main = "Gestation vs Residuals")

```
</br>
</br>
For the next iteration of the model, we added a quadratic term for gestation. While the quadratic term is statistically significant, the new model still displays the original downwards trend in the residual vs gestational period graph. Perhaps another transformation on gestation would be helpful -- a cubic term can be added. The other residuals plots also retain their trends from model 1.

### Model 3: Cubic Term Added
```{r, fig.width=5, fig.height=4}
model3 = lm(data = births_excl, BWTG ~ SEX + GEST + GEST2 + GEST3 + PARITY_truncated + PLUR_truncated + smoking_type + MAGE + MRACER + mortality, na.action = "na.exclude")
summary(model3)
par(mfrow = c(2,2))
plot(model3)
# plot(model3$fitted.values, model3$residuals)
plot(births_excl$GEST, model3$residuals, main = "Gestation vs Residuals")
```
</br>
</br>
After adding a cubic term to gestation, the residual vs gestational period shows a much more random pattern than before and the cubic term is statistically significant. It is worth investigating if adding a quartic term would help. The other residuals plots also retain their trends from model 1.


### Model 4: Quartic Term Added
```{r, fig.width=7, fig.height=5}
model4 = lm(data = births_excl, BWTG ~ SEX + GEST + GEST2 + GEST3 + GEST4 + PARITY_truncated + PLUR_truncated + smoking_type + MAGE + MRACER + mortality, na.action = "na.exclude")
summary(model4)
par(mfrow = c(2,2))
plot(model4)
# plot(model4$fitted.values, model4$residuals)
plot(births_excl$GEST, model4$residuals, main = "Gestation vs Residuals")
```
</br>
</br>
The addition of the quartic term does seem to help a bit, and it seems that we've gone as far as we can go with polynomial terms on gestation. Despite this, our assumption of normality still does not appear to be totally met, as shown by the Q-Q plot, and the residuals are not quite homoskedastic, as the variance of the residuals increases as the fitted values gets larger. Because of this, we should look into more robust methods. For that we will utilize M-estimation.

### Robust on Model 4

```{r, fig.width=7, fig.height=5}
robust1 <- rlm(data = births_excl, BWTG ~ SEX + GEST + GEST2 + GEST3 + GEST4 + PARITY_truncated + PLUR_truncated + smoking_type + MAGE + MRACER + mortality, na.action = "na.exclude")
summary(robust1)
par(mfrow = c(2,2))
plot(robust1)
# plot(robust1$fitted.values, robust1$residuals)
plot(births_excl$GEST, robust1$residuals, main = "Gestation vs Residuals")

#Check weights
robust1_weights = data.frame(bwt = births_excl$BWTG, gest = births_excl$GEST,
 resid=robust1$resid, weight=robust1$w)
robust1_weights[order(robust1$w)[c(1:5, (length(robust1$w)-5):length(robust1$w))],]
```
</br>
</br>

Our robust method downweights the weighting on observations with the largest residuals, as shown above. Looking at the output, it appears we still run into the same challenges as before with normality, even with this more robust approach. We will compare the approaches below by MSE using cross validation and splitting the data into test and training sets, but it is evident that an even more robust approach, such as quantile regression will be useful in exploring these relationships further.

### Cross Validation of Model
```{r}
births_cv<-births_excl[sample(nrow(births_excl)),]
folds<-cut(seq(1,nrow(births_cv)),breaks=10,labels=FALSE)
test_list<-list()
train_list<-list()
for(i in 1:10){
  test_indices<-which(folds==i,arr.ind=TRUE)
  births_test<-births_cv[test_indices,]
  test_list[[i]]<-births_test
  births_train<-births_cv[-test_indices,]
  train_list[[i]]<-births_train
}

#Train and test model1
model1_test_mse<-c()
for(i in 1:10){
  model1_train<-lm(data=train_list[[i]],BWTG~SEX+GEST+PARITY_truncated+PLUR_truncated+smoking_type+MAGE+MRACER+mortality,na.action="na.exclude")
  model1_test<-predict(model1_train,train_list[[i]])
  model1_test_mse[[i]]<-(mean((train_list[[i]]$BWTG-model1_test)^2))
}
test_mse<-c(mean(model1_test_mse))

#Train and test model2
model2_test_mse<-c()
for(i in 1:10){
  model2_train<-lm(data=train_list[[i]],BWTG~SEX+GEST+GEST2+PARITY_truncated+PLUR_truncated+smoking_type+MAGE+MRACER+mortality,na.action="na.exclude")
  model2_test<-predict(model2_train,train_list[[i]])
  model2_test_mse[[i]]<-mean((train_list[[i]]$BWTG-model2_test)^2)
}
test_mse<-append(test_mse, mean(model2_test_mse))

#Train and test model3
model3_test_mse<-c()
for(i in 1:10){
  model3_train<-lm(data=train_list[[i]],BWTG~SEX+GEST+GEST2+GEST3+PARITY_truncated+PLUR_truncated+smoking_type+MAGE+MRACER+mortality,na.action="na.exclude")
  model3_test<-predict(model3_train,train_list[[i]])
  model3_test_mse[[i]]<-mean((train_list[[i]]$BWTG-model3_test)^2)
}
test_mse<-append(test_mse, mean(model3_test_mse))

#Train and test model4
model4_test_mse<-c()
for(i in 1:10){
  model4_train<-lm(data=train_list[[i]],BWTG~SEX+GEST+GEST2+GEST3+GEST4+PARITY_truncated+PLUR_truncated+smoking_type+MAGE+MRACER+mortality,na.action="na.exclude")
  model4_test<-predict(model4_train,train_list[[i]])
  model4_test_mse[[i]]<-mean((train_list[[i]]$BWTG-model4_test)^2)
}
test_mse<-append(test_mse, mean(model4_test_mse))

robust1_test_mse<-c()
for(i in 1:10){
  robust1_train<-rlm(data=train_list[[i]],BWTG~SEX+GEST+GEST2+GEST3+GEST4+PARITY_truncated+PLUR_truncated+smoking_type+MAGE+MRACER+mortality,na.action="na.exclude")
  robust1_test<-predict(robust1_train,train_list[[i]])
  robust1_test_mse[[i]]<-mean((train_list[[i]]$BWTG-robust1_test)^2)
}
test_mse<-append(test_mse, mean(robust1_test_mse))

#Results
results_cv<-matrix(test_mse,ncol=5)
colnames(results_cv)<-c('model1','model2','model3','model4','robust1')
rownames(results_cv)<-c('Average MSE')
results<-as.table(results_cv)
results
```
We employed K-fold cross-validation to test the models on their ability to accurately predict. Following the technique, we shuffled the data to randomize the sample, and then partitioned the data into K=10 folds (groups). We used the 10 folds to form 10 pairs of test and training sets; every test set is composed a differnet fold of the 10 total folds, and each training set is the remaining nine folds combined. We used the training sets to produce the estimates from the model, and use the test sets to produce an error score, the MSE. We averaged the 10 MSEs for all for 5 the models we developed and used this output as our metric of comparison.

We prefer using these MSE scores over the in-sample results reported in our initial model estimates. Using cross validation allows us to mitigate the adverse effects of overfitting our model by adding more covariates. As illustrated by the table above, model4 has the lowest MSE out of all models utilised. This implies that model4 yields the least erroneous out of sample (test data) results, and thus indicates the least amount of overfitting. Nonetheless, MSE is decreasing across our models as we increase the number of covariates and the MSE produced by cross-validation of model4 is not substantially lower than the lower-degree models. Therefore, although our preferred model exhibits the least amount of overfitting across all models we tested, we should explore further functional forms to see if we can noticeably decrease MSE. Notaby, the robust model had the second lowest MSE. The robust model reduces the influence of outlying data, therefore we will consider also solving for the MAE across our models as it may increase the interpretability of the robust error. This indicates that by attempting to reduce the influence of outlying data in our model, we may have begun to overfit the data.

## Interpretations and Conclusions
```{r}
summary(model4)
```

Since our Model 4 seemed to fit the data best, we'll report our conclusions using this model. It's output is replicated above. 
</br>
</br>
Overall, the model does a good job of explaining birthweight, with our selected variables accounting for about 55% of the variation in birthweight in this sample, and each of the variables was highly predictive of birthweight.
</br>
</br>
**Sex** - On average male babies are heavier by about 125 grams, holding all else constant
</br>
</br>
**Gestation** - Given that there is a polynomial relationship between gestation and birthweight, it is hard to interpret this relationship simply, but in general, longer gestation periods are expected to yield heavier babies.
</br>
</br>
**Parity** - In general, children with a higher parity, that is to say children with more older siblings, tend to be heavier than first-borns. Second-borns are expected to be 85 grams heavier than the average first born, holding all else constant. Other higher-parity children are expected to be even heavier, by about 100 grams versus a first-born.
</br>
</br>
**Plurality** - Twins and triplets (or more) are expected to be much lighter than single-birth babies, by about 310-320 grams.
</br>
</br>
**Smoking** - smoking is expected to have a negative effect on babies' birthweights. Mothers who smoked only during pregnancy are expected to have babies which are about 160 grams lighter. Mothers who smoked in the three months leading up to pregnancy as well are expected to have babies which are about 200 grams lighter. Finally, even mothers who only smoked leading up to pregnancy, but not during, also are expected to have lighter babies, but only by about 10 grams.
</br>
</br>
**Mother's Age** - Our model indicates a small but statistically significant positive effect of mother's age on birthweight. For every year a mother gets older, her baby is expected to be about 4 grams heavier.
</br>
</br>
**Race** - Race is expected to have a significant impact on birthweight. Versus the other-nonwhite category, white mothers are expected to have babies that are about 80 grams heavier. This effect varies by race, but the race which is associated with the lightest babies is Japanese, at about 115 grams lighter than the other category, and the race associated with the heaviest babies is Hawaiian, at about 200 grams heavier.
</br>
</br>
**Infant Mortality Rates** - We used infant mortality rates in county of birth as a proxy for socioeonomic status, which turned out to be a significant predictor of birthweight. However, it's practical impact is not very large. On average, each one additional percentage point increase in infant mortality rate of a county is associated with an increase of about 20 grams in birthweight. It is important to note that the entire range of infant mortality rates in the counties our data comes from is only about 1.5 percentage points, so the practical impact of infant mortality rates on birthweight is not large.








