---
title: "440 Case Study I"
author: "Jake Epstein, Daniel Spottiswood, Michael Tan, Sahil Patel, Man-Lin Hsiao"
date: "9/3/2019"
output: html_document
---

## Set Up, Load, Clean Data
```{r}
## load packages
library(dplyr)
library(ggplot2)

## read in data
births = read.csv("data/Yr1116Birth.csv", na.strings = "9999")
deaths = read.csv("data/Yr1116Death.csv")
 
## rewrite NAs
births$SEX[which(births$SEX == 9)] = NA
births$CIGPN[which(births$CIGPN == 99)] = NA
births$CIGFN[which(births$CIGFN == 99)] = NA
births$CIGSN[which(births$CIGSN == 99)] = NA
births$CIGLN[which(births$CIGLN == 99)] = NA
births$PARITY[which(births$PARITY == 99)] = NA
births$PLUR[which(births$PLUR == 99)] = NA
births$GEST[which(births$GEST == 99)] = NA
births$MAGE[which(births$MAGE == 99)] = NA
```


## Exploratory Data Analysis

### Birthweight

### Smoking

### Parity

```{r}
# Check the parity frequencies
births$PARITY = as.factor(births$PARITY)
ggplot(data = births, mapping = aes(x = PARITY, y = BWTG)) +
  geom_boxplot() + xlab("Parity") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + theme_minimal()
ggplot(births, aes(x = PARITY)) +
        stat_count()

# Model without any change, significant up to the low teens
parity_reg = lm(BWTG ~ PARITY, births)
summary(parity_reg)

# Create variable 5+ and rerun diagnostics
births$PARITY = factor(births$PARITY, levels=c(levels(births$PARITY), "5+"))
births$PARITY[which(as.integer(births$PARITY) >= 5)] = "5+"
ggplot(births, aes(x = PARITY)) +
        stat_count()
ggplot(data = births, mapping = aes(x = PARITY, y = BWTG)) +
  geom_boxplot() + xlab("Parity") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016")
```

```{r}
parity_reg = lm(BWTG ~ PARITY + mortality + GEST, births)
summary(parity_reg)
```
We see that parity does have a significant impact on birth weight. Independent of other variables, the second child has the highest expected birth weight and then the expected birth weight slightly decreases for later children. Specifically, the sixth child has a lower expected birth weight than the first child. We chose to group parities of five or more as it provides more concise interpretation and prevents overfitting for parities that have very few data points. We see that the frequency of parity seems to decrease in an exponential fashion.

We need to decide if we want this to be a categorical variable?? If not I will redo and probably try to fit a quadratic.

### Plurality

```{r}
# Check the plurality frequencies
births$PLUR = as.factor(births$PLUR)
ggplot(data = births, mapping = aes(x = PLUR, y = BWTG)) +
  geom_boxplot() + xlab("Plurality") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + theme_minimal()
ggplot(births, aes(x = PLUR)) +
        stat_count()

plurality_reg = lm(BWTG ~ PLUR, births)
summary(plurality_reg)
```

We see that the frequency of higher pluralities decrease rapidly. The results; however, are still statistically significant and we see that higher plurality is linked with lower birth weights.

### Gestation

### Age of Mother

```{r}
# Check the plurality frequencies
ggplot(data = births, mapping = aes(x = MAGE, y = BWTG)) + xlab("Age of Mother (years)") + ylab("Birth weight (g)") + 
  ggtitle("NC Births, 2011-2016") + theme_minimal()
ggplot(births, aes(x = MAGE)) +
        stat_count()

mage_reg = lm(BWTG ~ MAGE, births)
summary(mage_reg)
```
Mother's age seems to be fairly normally distributed with a mean in the late twenties.

### Race of Mother

### County / Socioeconomic Status
```{r}
#calculate infant mortality by county
deaths_by_county = deaths %>%
  # filter(!is.na(cores)) %>%
  group_by(cores) %>%
  summarize(n_deaths = n()) %>%
  rename(CORES=cores)

births_by_county = births %>%
  # filter(!is.na(CORES)) %>%
  group_by(CORES) %>%
  summarize(n_births = n())

infant_mortality = merge(deaths_by_county, births_by_county, by = "CORES") %>%
  mutate(mortality = n_deaths/n_births) %>%
  select(-n_births, -n_deaths)

births = merge(births, infant_mortality, by = "CORES")


```
